Temporal ordering and registration of images in studies of developmental dynamics

### 2014-08-30 image_analysis_paper2/SI_text.tex
small edits to paper

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij}).
\end{equation}
%
%
Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
%\begin{equation}
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and
\begin{equation} \label{eq:H_low_rank}
	H \approx
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
%
Then,
\begin{equation} \label{eq:R_hat}
\hat{R} =
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & |
\end{bmatrix},
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
%
To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix,
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T,
\end{equation}
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
%
We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
discrete shift.

@@

\begin{table}[t]
%\caption{Outline of algorithm used to temporally order images}
\setlength{\fboxrule}{2pt}
\framebox[\columnwidth]{
\begin{minipage}{0.9\columnwidth}
{\bf Outline of algorithm used to register and order images}
\begin{enumerate}
\item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
Retain both the optimal pairwise rotations as well as the minimum distances obtained when pairs are optimally aligned.
%
\item Compute the corresponding rotation matrices $R(\theta_{ij})$ from \eqref{eq:R_theta}.
%
\item Select $\epsilon$ for use in the diffusion maps kernel. 
%
Taking $\epsilon$ to be on the order of the median of the pairwise distances often yields good results. 
\item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\theta_{ij})$ are computed in step 2.
%
\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{3m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{3m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{3m}|$.
%
\item Estimate the optimal three-dimensional rotation for each image: stack the first three eigenvectors $\phi_1, \phi_2, \phi_3$ into a $3m \times 3$ matrix, and then divide this matrix into $m$ $3 \times 3$ blocks, $\hat{R}_1, \dots, \hat{R}_m$. Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}. If most of the rotations are improper ($det(R) = -1$), flip the sign of one of the eigenvectors.
\item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
\item For each image $i$, compute the optimal rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding angle using \eqref{eq:R_theta}. 
\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $3 \le k \le 4$.
\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,1}(i)$.
\end{enumerate} 
\end{minipage}
}
\end{table}

@@

We can use the distribution of the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
%
In diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector, and $|\lambda_k|$ gives a measure of the importance of coordinate $\phi_k$. 
%
Therefore, we expect to see a ``spectral gap'' in the eigenvalues which separates the meaningful coordinates from those corresponding to noise.
%
However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}.
%
Therefore, one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.

@@

\begin{figure}
\includegraphics{figS1}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green). The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position(s) where the cross-section of an embryo is imaged. (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. Images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (right arrow in the left image). }
\label{fig:ap_dv}
\end{figure}

@@

\begin{figure*}	
\includegraphics{figS6}
\caption{Fixed embryo imaging data set. {\it (A)} Fixed images used for the results in \fig~4. The images have been preprocessed as described above. {\it (B)} Data set from A after registration and ordering. The expert rank for each image is indicated. }
\end{figure*}

>>>>

Let $d$ denote the dimension of the rotation matrices we are considering (for planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks, with the $i,j^{th}$ block of $H$, $H_{ij}$, defined as
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij}).
\end{equation}
%
%
Under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$
%\begin{equation}
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and
\begin{equation} \label{eq:H_low_rank}
	H \approx
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
It follows directly from \eqref{eq:H_low_rank} that the top block eigenvector of $H$ contains our best estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$ ordered so that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$.
%
Then,
\begin{equation} \label{eq:R_hat}
\hat{R} =
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & |
\end{bmatrix},
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is (nearly) the estimate for $R(\theta_i)$.
%
To obtain our estimate of $R(\theta_i$), denoted $R_{i, est}$, we project $\hat{R}_i$ onto the closest orthogonal matrix,
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T,
\end{equation}
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations 
(note that systematically incorporating improper rotations is also possible \cite{goemans1995improved, bandeira2013cheeger}).
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.
%
We note that, in our actual computations, the pairwise rotations $\theta_{ij}$ are computed in a discrete setting, then the overall
synchronization is performed in the continuum context to obtain $\theta_i$, and the results are rounded to give the closest
discrete shift.

@@

\begin{table}[t]
%\caption{Outline of algorithm used to temporally order images}
\setlength{\fboxrule}{2pt}
\framebox[\columnwidth]{
\begin{minipage}{0.9\columnwidth}
{\bf Outline of algorithm used to register and order images}
\begin{enumerate}
\item Compute the alignments for each pair of images, as described in \eqref{eq:opt_pairwise}.
Retain both the optimal pairwise rotations as well as the minimum distances obtained when pairs are optimally aligned.
%
\item Compute the corresponding rotation matrices $R(\theta_{ij})$ from \eqref{eq:R_theta}.
%
\item Select $\epsilon$ for use in the diffusion maps kernel. 
%
Taking $\epsilon$ to be on the order of the median of the pairwise distances often yields good results. 
\item Compute the matrix $S$ in \eqref{eq:vdm_S}. 
$A_{ij}$ is computed from \eqref{eq:dmaps_W} and \eqref{eq:dmaps_A}, using the pairwise distances from step 1, and $H_{ij} = R(\theta_{ij})$ are computed in step 2.
%
\item Compute the eigenvalues  $\lambda_1, \lambda_2, \dots, \lambda_{2m}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{2m}$ of $S$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{2m}|$.
%
\item Estimate the optimal three-dimensional rotation for each image: stack the first two eigenvectors $\phi_1, \phi_2$ into a $2m \times 2$ matrix, and then divide this matrix into $m$ $2 \times 2$ blocks, $\hat{R}_1, \dots, \hat{R}_m$. Compute the estimated rotations for each of these $m$ blocks as in \eqref{eq:R_est}. If most of the rotations are improper ($det(R) = -1$), flip the sign of one of the eigenvectors.
\item Multiply each estimated rotation $R_{i, est}$ by $R_{1, est}^T$ (so that $R_{1, est}$ will become the identity matrix).
\item For each image $i$, compute the optimal rotation to align the image by converting $R_{i, est}^T$ (note the transpose) to the corresponding angle using \eqref{eq:R_theta}. 
\item Compute the first embedding coordinate $\psi_{k,1}(i)$ as described in \eqref{eq:vdm_coord} and \eqref{eq:first_embed_vdm}, where $3 \le k \le 4$.
\item To order the images, sort them by the values of this embedding coordinate $\psi_{k,l}(i)$.
\end{enumerate} 
\end{minipage}
}
\end{table}

@@

We can use the eigenvalues from (vector) diffusion maps to help deduce the dimensionality of the data.
%
In diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector, and $|\lambda_k|$ gives a measure of the importance of coordinate $\phi_k$. 
%
We therefore expect to see a ``spectral gap'' in the eigenvalues which separates the meaningful coordinates from those corresponding to noise.
%
However, some embedding coordinates which appear meaningful according to the eigenvalues may be harmonics of previous coordinates \cite{ferguson2010systematic}, 
and one must visually check for correlations among potential embedding coordinates before deducing the true dimensionality of the data.

@@

\begin{figure}
\includegraphics{figS1}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Twi (green). The embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position where the cross-section of an embryo is imaged. (Right) A dorsoventral view of the cross-section of the {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. Images were collected at the focal plane $\sim 18\%$ from the posterior pole of an embryo (arrow in the left image). }
\label{fig:ap_dv}
\end{figure}

@@

\begin{figure*}	
\includegraphics{figS6}
\caption{Fixed embryo imaging data set. {\it (A)} Fixed images used for the results in \fig~4. The images have been preprocessed as described above. {\it (B)} Data set from A after registration and ordering. The rank provided by a trained embryologist for each image is indicated. }
\end{figure*}
